A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-04 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: another pass at classical probability

\textbf{Choquet theory.} Having reviewed the use of probability measures in classical mechanics, we can ask the more general question: when can we represent ensembles in a generic ensemble space as a probability measure? This question is treated by Choquet theory, though in not the full generality that we need it. Let us therefore review the main results of Choquet theory, before discussing how it connects to ensembles spaces.

Mathematically, the goal of Choquet theory is to represent a convex set as a probability measure over the extreme points. Let $E$ be a convex subset of a locally convex space $V$. We say that $\ens \in E$ is represented by a probability measure $\mu : \Sigma_{E} \to [0,1]$ if $f(\ens) = \int_E f d\mu$ for every continuous linear functional $f$ on $V$. The Choquet theorem states:
\begin{thrm}[Choquet]
	Let $E$ be a metrizable compact convex subset of a locally convex space $V$, and let $\ens \in E$. Then there exists a probability measure $\mu_{\ens}$ on $X$ which represents $\ens$ and is supported by the extreme points of $X$.
\end{thrm}
Since we saw that an ensemble space is convex subset of a locally convex topological vector space, the relevance of Choquet theory is straight forward.

% Physical meaning of representation with a measure.
First, let us understand the physical significance. An ensemble $\ens$ can be seen as a probability measure $\mu$ over the ``pure states'' (i.e. the extreme points of $E$) if the value for every statistical quantity $f$ for $\ens$ can be recovered as an expectation over the pure states using the probability measure. That is, all statistical properties of $\ens$ can be recovered by the probability measure. This is what allows to think go $\ens$ as a probability distribution. Note, however, that the Choquet theorem does not guarantee uniqueness. The probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choque simplex by the property of having a single decomposition in terms of extreme points.

% Limitation of applying Choquet theory to our case
We should now look at the requirements of the theorem, to see whether we can always apply it to an ensemble space. First of all, and ensemble space $\Ens$ is a convex subset of a convex sapce $V$. Moreover, since every Hausdorff first-countable topological vector space is metrizable, $\Ens$ is metrizable. However, $\Ens$ is not necessarily compact. In fact, $\Ens$ is not necessarily a closed subset of $V$. Consider the space of probability measures over $\mathbb{R}$. Since $\mathbb{R}$ is not compact, the space of probability measure over $\mathbb{R}$ cannot be compact.\footnote{TODO: Check with mathematicians} Moreover, the set of extreme points is given by the Dirac measures (i.e. probability one over a single point), which would have minus infinite entropy, and therefore cannot be part of the ensemble space. Therefore the theorem does not apply in general.

% Define "Choquet space" as those spaces where we can apply Choquet theory
We can apply Choquet theorem in the case where the closure of $\Ens$ in $V$ is a compact set $E$, which will also be convex\footnote{The closure of a convex set is convex. TODO: finalize the proof} and metrizable. In this case, we are guaranteed that all ensembles allows a representation through a probability on points that may corresponds idealized ensembles (i.e. the Dirac measure) that are not, even conceptually, realizable. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distribution over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems.

Note that ensembles over classical spaces will be represented by a unique probability measure, therefore the corresponding closure of the ensemble space is a Choquet simplex. However, quantum mixed states can be decomposed in multiple ways in terms of pure states, and therefore the closure of a quantum ensemble space is not a simplex.

@@

Chouqet theory tells, in the realm where it can be applied, that classical probability fails in quantum mechanics precisely because the ensemble space of a quantum system allows multiple decomposition.

@@

>>>>

\textbf{Choquet theory.} Having reviewed the use of probability measures in classical mechanics, we can ask the more general question: when can we represent ensembles in a generic ensemble space as a probability measure? A very similar question is the subject of Choquet theory.

First, we have to define what it means to represent an ensemble with a probability measure. As a first step, let $E$ be a convex subset of a locally convex space $V$. We say that $\ens \in E$ is represented by a probability measure $p : \Sigma_{E} \to [0,1]$ if $f(\ens) = \int_E f dp$ for every continuous linear functional $f$ on $V$. Physically, instead of describing our ensemble $\ens$ as a point in our ensemble space identified by the value of all possible statistical variables $f(\ens)$, we represent it as a probability measure $p$ and all the associated expectation values.

In this first step, we are using a probability measure on $E$, so we are defining a probability distribution over ensembles. As a second step, we want the probability measure to be defined over the possible cases, the sample space. To do that, Choquet theory identifies each possible case with the Dirac measure associated to that case, which is an extreme point of the convex set $E$. The result is the following:
\begin{thrm}[Choquet]
	Let $E$ be a metrizable compact convex subset of a locally convex space $V$, and let $\ens \in E$. Then there exists a probability measure $\mu_{\ens}$ on $X$ which represents $\ens$ and is supported by the extreme points of $X$.
\end{thrm}

Note that while an ensemble space $\Ens$ is a metrizable convex subset of a locally convex TVS $V$, it is not necessarily compact. As we saw, the Dirac measures should not be considered valid ensembles in the continuum case. Therefore Choquet theory cannot be applied to our general case. When we are drawing the parallel, we will implicitly assume that $E$ is a compact set that is also the closure of $\Ens$ in an appropriate convex TVS $V$. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distribution over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems. When Choquet theory applies, then, we are able to always characterize an ensemble in terms of a probability distributions over its, maybe idealized, most refined ensembles.

The previous theorem does not, however, guarantee uniqueness. The above theorem, in fact, applies in the quantum case. The Bloch ball, in fact, is a compact convex subset and any point can be seen as a convex combination of points on the surface. The center, which represent the maximally mixed state, can be thought as an equal mixture of any two opposite points, or a uniform distribution over the whole surface. The fact that any ensemble can be seen as a mixture of a set of pure states, then, is not a property of classical spaces. The property of classical spaces is that this representation is unique.

Mathematically, the probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choquet simplex by the property of having a single decomposition in terms of extreme points. The classical case, that is when the ensemble space coincide with the space of all probability measures over a set of states, is a Choquet simplex.

@@

The core insight from Chouqet theory, then, is that both in the classical and quantum case we can represent ensembles as a probability distribution over a set of states. However, only in the classical case this representation is unique. In the quantum case, mixed states can be decomposed into pure states in multiple ways, and it is this multiple decomposition that makes the ensemble space not a simplex, and therefore the space not classical. Unique representation of ensembles is a desirable property, and standard probability measures are not going to provide it in general.

@@

As a comparison, in the previous case all probability measures were given over a uniquely defined set of points, the extreme points of the ensemble space, but the probability measure may not be unique. In this case, the pseudo-probability associated to each ensemble is unique, but the space over which the pseudo-probability is defined is not.
